Properties of Winning Iterated Prisoner's Dilemma Strategies.

### 2024-11-04 paper/main.tex
first draft of revision

%%%%Packages%%%%
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{graphics}
\usepackage{multicol}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{setspace}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{hyperref}
\usepackage{color,colortbl}
\usepackage{array}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{wrapfig, blindtext}
\usepackage{soul}
\usepackage{enumerate}
\usepackage[table]{xcolor}
\usepackage{authblk}
\usepackage{listings}
%%%%%%%%%%%%%%%%%

@@

We believe that the discussion on the properties of winning strategies holds
significant importance. It aims to provide guidance to researchers designing new
strategies and those training strategies. Specifically, much like the recognized
value of diversity in training datasets, such as variations in image perspective,
skin color, etc., are critical in training accurate and generalizable machine
learning models, we show that diversity in the population of opponent strategies
is of paramount importance in the construction and evaluation of game theory
strategies. Moreover, conducting a similar
analysis can shed light on already trained strategies, aiding in understanding
the key features they have autonomously developed during their training
processes.

@@

To run a tournament, only a few lines of code are required
(Figure~\ref{fig:axelrod_code}). Specifically, one needs to define the list of
strategies that the players use when participating in the tournament, the number
of repetitions, the number of turns or the probability of each match ending, and
the probability of noise. We demonstrate two examples in
Figure~\ref{fig:axelrod_code}: one with a fixed number of turns and one with a
probabilistic ending. A tournament is an instance of the `Tournament` class. To
execute the tournament, users need to run the `play()` method, which returns an
instance of the `Results` class. This instance includes many details of the
tournament, such as the winner and the average score. Additionally, a more
detailed summary of the tournament is provided, which we use in this analysis
and describe in detail below.

@@

>>> results = prob_noisy_end_tournament.play()
\end{lstlisting}
\caption{{\bf Example usage of the \axelrod package for running tournaments.}
The strategies that players use are saved in a list, which is then passed to the
Tournament class. In this example, strategies we previously discussed, such as \TFT, Generous Tit-for-Tat, Gradual, and a stochastic strategy
evolved through reinforcement learning, are included. We create a standard
tournament where noise is set to 0, as
well as a noisy tournament with probabilistic ending. Once an instance of the
tournament class is defined, executing the tournament is straightforward. The play()
method generates all possible pairs from the list of strategies, including
pairs where each strategy plays against itself, and then iterates through each
match to play it. Each match is repeated according to the specified number of
repetitions, with results aggregated to a tournament summary.}\label{fig:axelrod_code}
\end{figure}

@@

The summary contains statistics regarding each strategy that participated in the
tournament, such as its rank, cooperation rate, time spent in each state, and
the probability of cooperating after each of the four possible outcomes of the
previous round. In our analysis, we will use the measures for each strategy that
are provided in the summary, as well as additional measures we calculated.
Namely, these include the SSE error, the average cooperation in each tournament,
the maximum, and the minimum values. The SSE (introduced in~\cite{Knight2024})
shows how close a strategy is to behaving as a ZDs, and subsequently, in an
extortionate way. We also consider how each strategy's cooperation rate \(C_r\)
compares to those of the tournament as a whole, for example, by comparing
\(C_r\) to \(C_{\text{max}}\).

@@

\newcolumntype{g}{>{\columncolor{Gray}}c}
\begin{table}[t]
    \begin{center}
    \resizebox{\textwidth}{!}{
    \begin{tabular}{ccccccgcgcgcgcg}
    \toprule
    & & & & & &   \multicolumn{8}{g}{Rates}  \\
    Rank & Name & Median score & Cooperation rating $(C_r)$ & Win & Initial C &
    CC & CD & DC & DD & CC to C & CD to C & DC to C & DD to C \\
    0 &  EvolvedLookerUp2 2 2 & 2.97 & 0.705 & 28.0 & 1.0 & 0.639 & 0.066 & 0.189 &
    0.106 & 0.836 & 0.481 & 0.568 & 0.8 \\
    1 &  Evolved FSM 16 Noise 05 & 2.875 & 0.697 & 21.0 & 1.0 & 0.676 &
    0.020 & 0.135 & 0.168 & 0.985 & 0.571 & 0.392 & 0.07 \\
    2 & PSO Gambler 1 1 1 & 2.874 & 0.684 &  23.0 &     1.0 &    0.651 &    0.034 &    0.152 &    0.164
    & 1.000 & 0.283 & 0.000 & 0.136 \\
    3 &  PSO Gambler Mem1 &  2.861 &        0.706 &  23.0 &      1.0 &    0.663
    &    0.042 &    0.145 &    0.150 &  1.000 &  0.510 &  0.000 &  0.122 \\
    4 &          Winner12 &  2.835 &        0.682 &  20.0 &      1.0 &
    0.651 &    0.031 &    0.141 &    0.177 &  1.000 &  0.441 &  0.000 &  0.462 \\
    $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ &
    $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ \\
    \bottomrule
    \end{tabular}}
\end{center}
\caption{\textbf{Result Summary Example of a Tournament.}
A result summary consists of \(N\) rows, with each row containing information
for each strategy that participated in the tournament. This information includes
the strategy's rank (\(R\)), median score, the cooperation rate (\(C_r\)), the number of
match wins, and the probability that the strategy cooperated in the opening
move. Additionally, it provides the probabilities of a strategy being in any of
the four states ($CC, CD, DC, DD$) and the cooperation rate after each state.}\label{table:output_result}
\end{table}

@@

\newcolumntype{g}{>{\columncolor{Gray}}c}
\begin{table}[!htbp]
    \begin{center}
    \resizebox{.8\textwidth}{!}{
        \input{correlations_table.tex}
    }
\end{center}
\caption{\textbf{Correlations between the features of Table~\ref{table:manual_features}
and the normalised rank and the median score.}
For each type of tournament, standard, noisy, probabilistic ending, and noisy
probabilistic ending, we run a correlation analsysif. For each tournament we
check the correlation between each feature used in our analysis and the normalised
random and median score. Note that the correlation coefficients are calculated
using the Spearman's rank correlation coefficient. A negative value
indicates a negative correlation and in the case of the normalised rank
a smaller rank translats to better position in the tournament.}\label{table:correlations}
\end{table}

@@

This manuscript explores the performance of \numberofstrategies strategies in
the \IPD{} in thousands of computer tournaments. The collection of computer
tournaments presented here is the largest and most diverse in the literature.
The \numberofstrategies strategies are drawn from \axelrod{} library and include
strategies from the \IPD{} literature. The computer tournaments encompass four
different types. So, what is the best way to play the \IPD{}? And is there a
single dominant strategy for the \IPD{}? There was not a single strategy within
the collection of \numberofstrategies strategies that managed to perform well in
all the tournament variations it competed in. A strategy ranking highly in a
specific environment did not guarantee its success over different tournament
types, with a few exceptions - strategies that generalize better. Already
well-known in the AI/ML literature, adding noise to training data leads to more
robust models. We see that clearly here, where the strategies trained for noise
(or designed for noise) tend to be better generalists. There were instances
where a few strategies trained in narrow conditions outperformed more generalist
strategies, as they tend to overfit. However, the strategies trained with noise
perform well in general, whilst the strategies trained specifically on no noise
or small subpopulations do not.

>>>>

%%%%Packages%%%%
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{graphics}
\usepackage{multicol}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{setspace}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{hyperref}
\usepackage{color,colortbl}
\usepackage{array}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{wrapfig, blindtext}
\usepackage{soul}
\usepackage{enumerate}
\usepackage[table]{xcolor}
\usepackage{authblk}
\usepackage{listings}
\usepackage{xparse}

\NewDocumentCommand{\codeword}{v}{%
\texttt{\textcolor{teal}{#1}}%
}

%%%%%%%%%%%%%%%%%

@@

We believe that the discussion on the properties of winning strategies holds
significant importance. It aims to provide guidance to researchers designing new
strategies and those training strategies. Specifically, much like the recognized
value of diversity in training datasets~\cite{Gong:IEEE:2019}, such as
variations in image perspective, skin color, etc., are critical in training
accurate and generalizable machine learning models, we show that diversity in
the population of opponent strategies is of paramount importance in the
construction and evaluation of game theory strategies. Moreover, conducting a
similar analysis can shed light on already trained strategies, aiding in
understanding the key features they have autonomously developed during their
training processes.

@@

To run a tournament, only a few lines of code are required
(Figure~\ref{fig:axelrod_code}). Specifically, one needs to define the list of
strategies that the players use when participating in the tournament, the number
of repetitions, the number of turns or the probability of each match ending, and
the probability of noise. We demonstrate two examples in
Figure~\ref{fig:axelrod_code}: one with a fixed number of turns and one with a
probabilistic ending. A tournament is an instance of the \texttt{Tournament}
class. To execute the tournament, users need to run the \texttt{play()}
method, which returns an instance of the \texttt{ResultSet} class. This
instance includes many details of the tournament, such as the winner and the
average score of the participants. Additionally, the instance contains a more
detailed summary of the tournament, which we use in our analysis. We describe
the results summary in detail below.

@@

>>> results = prob_noisy_end_tournament.play()
\end{lstlisting}
\caption{{\bf Example usage of the \axelrod{} package for running tournaments.}
The strategies that players use are saved in a list, which is then passed to the
Tournament class. In this example, strategies we previously discussed, such as \TFT, Generous Tit-for-Tat, Gradual, and a stochastic strategy
evolved through reinforcement learning, are included. We create a standard
tournament where \texttt{noise} is set to 0, as
well as a noisy tournament with probabilistic ending. Once an instance of the
tournament class is defined, executing the tournament is straightforward. The \texttt{play()}
method generates all possible pairs from the list of strategies, including
pairs where each strategy plays against itself, and then iterates through each
match to play it. Each match is repeated according to the specified number of
repetitions, with results aggregated to a tournament summary.}\label{fig:axelrod_code}
\end{figure}

@@

The summary contains statistics regarding each strategy that participated in the
tournament, such as its rank, cooperation rate, time spent in each state when
only a single past round is considered, and the probability of cooperating after
each of the four possible outcomes of the previous round. In our analysis, we
will use the measures for each strategy provided in the summary, as well as
additional measures we calculated. Namely, these include the SSE error, the
average, median, maximum, and minimum cooperation rates in each tournament. The
SSE (introduced in~\cite{Knight2024}) shows how closely a strategy behaves as a
zero-determinant strategy and subsequently in an extortionate way. We also
consider how each strategy's cooperation rate \(C_r\) compares to those of the
tournament as a whole, for example, by comparing \(C_r\) to \(C_{\text{max}}\).

\newcolumntype{g}{>{\columncolor{Gray}}c}
\begin{table}[b]
    \begin{center}
    \resizebox{\textwidth}{!}{
    \begin{tabular}{ccccccgcgcgcgcg}
    \toprule
    & & & & & &   \multicolumn{8}{g}{Rates}  \\
    Rank & Name & Median score & Cooperation rating $(C_r)$ & Win & Initial C &
    CC & CD & DC & DD & CC to C & CD to C & DC to C & DD to C \\
    0 &  EvolvedLookerUp2 2 2 & 2.97 & 0.705 & 28.0 & 1.0 & 0.639 & 0.066 & 0.189 &
    0.106 & 0.836 & 0.481 & 0.568 & 0.8 \\
    1 &  Evolved FSM 16 Noise 05 & 2.875 & 0.697 & 21.0 & 1.0 & 0.676 &
    0.020 & 0.135 & 0.168 & 0.985 & 0.571 & 0.392 & 0.07 \\
    2 & PSO Gambler 1 1 1 & 2.874 & 0.684 &  23.0 &     1.0 &    0.651 &    0.034 &    0.152 &    0.164
    & 1.000 & 0.283 & 0.000 & 0.136 \\
    3 &  PSO Gambler Mem1 &  2.861 &        0.706 &  23.0 &      1.0 &    0.663
    &    0.042 &    0.145 &    0.150 &  1.000 &  0.510 &  0.000 &  0.122 \\
    4 &          Winner12 &  2.835 &        0.682 &  20.0 &      1.0 &
    0.651 &    0.031 &    0.141 &    0.177 &  1.000 &  0.441 &  0.000 &  0.462 \\
    $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ &
    $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ \\
    \bottomrule
    \end{tabular}}
\end{center}
\caption{\textbf{Result Summary Example of a Tournament.}
A result summary consists of \(N\) rows, with each row containing information
for each strategy that participated in the tournament. This information includes
the strategy's rank (\(R\)), median score, the cooperation rate (\(C_r\)), the number of
match wins, and the probability that the strategy cooperated in the opening
move. Additionally, it provides the probabilities of a strategy being in any of
the four states ($CC, CD, DC, DD$) and the cooperation rate after each state.}\label{table:output_result}
\end{table}

@@

@@

\newcolumntype{g}{>{\columncolor{Gray}}c}
\begin{table}[!htbp]
    \begin{center}
    \resizebox{.8\textwidth}{!}{
        \input{correlations_table.tex}
    }
\end{center}
\caption{\textbf{Correlations between the features of Table~\ref{table:manual_features}
and the normalised rank and the median score.} For each type of
tournament, standard, noisy, probabilistic ending, and noisy probabilistic
ending, we conduct a correlation analysis. For each tournament, we check the
correlation between each feature used in our analysis and the normalized random
and median scores. Note that the correlation coefficients are calculated using
Spearman's rank correlation coefficient. A negative value indicates a negative
correlation, and in the case of the normalized rank, a smaller rank translates
to a better position in the tournament.}\label{table:correlations}
\end{table}

@@

This manuscript explores the performance of \numberofstrategies strategies in
the \IPD{} in thousands of computer tournaments. The collection of computer
tournaments presented here is the largest and most diverse in the literature.
The \numberofstrategies strategies are drawn from \axelrod{} library and include
strategies from the \IPD{} literature. The computer tournaments encompass four
different types. So, what is the best way to play the \IPD{}? And is there a
single dominant strategy for the \IPD{}? There was not a single strategy within
the collection of \numberofstrategies strategies that managed to perform well in
all the tournament variations it competed in. A strategy ranking highly in a
specific environment did not guarantee its success over different tournament
types, with a few exceptions - strategies that generalize better. Already
well-known in the AI/ML literature, adding noise to training data leads to more
robust models~\cite{Bishop1995}. We see that clearly here, where the strategies trained for noise
(or designed for noise) tend to be better generalists. There were instances
where a few strategies trained in narrow conditions outperformed more generalist
strategies, as they tend to overfit. However, the strategies trained with noise
perform well in general, whilst the strategies trained specifically on no noise
or small subpopulations do not.
